---
title: "Generate INAR(1) process"
date: "`r Sys.Date()`"
bibliography: bibliography.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{INAr package, Generate INAR(1) process}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeyword{PDF}
  %\VignetteKeyword{HTML}
---

# Load library

```{r warning=FALSE}
library(INAr)
library(ggplot2)

set.seed(6)
```

# Simulate an INAR(1) process

The following chunk generates a univariate INAR(1) time series using Poisson innovations.  
The thinning parameter \(a = 0.45\) controls the autocorrelation, while `burnout = 500` ensures the process reaches stationarity before recording the final 200 observations.

```{r warning=FALSE}
X <- genINAR(
  n       = 200,      # length of the series
  a       = 0.45,     # thinning parameter (autocorrelation)
  par     = 3,        # parameter of the innovation distribution
  inn     = "poi",    # Poisson innovations
  burnout = 500)      # burn-in to reach stationarity
```

# Visualise the simulated series

This final chunk constructs a simple data frame and plots the INAR(1) trajectory using a clean ggplot line graph.

```{r warning=FALSE}
df <- data.frame(time = 1:200, value = X)

ggplot(df, aes(time, value.X)) +
  geom_line(color = "darkorange", linewidth = 1) +
  theme_minimal() +
  labs(title = "Simulated INAR(1) Time Series",
       x = "Time", 
       y = "Counts")

```

